---
title: 'Metaheuristics and Optimization in R'
author: "DS 775"
fontsize: '12pt'
output:
  beamer_presentation:
    template: '../../beamer169.tex'
    colortheme: "seahorse"
    keep_tex: true
---

## This Unit

- Goal 1:  introduce three metaheuristic (optimization) techniques
    - Multistart methods
    - Simulated Annealing
    - Genetic Algorithms
    
- Goal 2:  introduce some optimization tools available in R including
    - optim() in the 'stats' package for local search
    - genetic algorithms in 'GA' and 'gramEvol' packages
    - multiple algorithms in NLopt accessible through 'nloptr'
    - simulated annealing through optim() and GenSA()
    
<div class="notes">
- goes with audio01.mp3

</div>

## Metaheuristics

- problems so complex that we cannot guarantee an optimal solution
    - complex nonlinearities leading to many potentially optimal points
    - combinatorial optimization with too many possible solutions
- Goal:  find a good feasible solutions
- algorithms are often iterative

<div class="notes">
goes with audio02.mp3
</div>

## Maximize vs Minimize

\includegraphics[width=4.0in]{picture1.pdf}

- from now on we'll talk about minimization

<div class="notes">
goes with audio03.mp3
</div>


## Local Search

- start with $x_0$ and evaluate $f(x_0)$
- choose $x_1$ close to $x_0$ so that hopefully $f(x_1)$ is smaller
- repeat until close enough to minimum

<div class="notes">
goes with audio04.mp3
</div>


## Local Search - Continuous Variables

\includegraphics[width=4.0in]{picture2.pdf}

- for smooth functions like this one we can use (approximate) calculus to descend the hill until we reach the bottom

<div class="notes">
goes with audio05.mp3
</div>

## Local Search - Rough Functions

\includegraphics[width=4.0in]{picture3.pdf}

- for rough functions calculus won't help but other search methods like a genetic algorithm or simulated annealing can be used

<div class="notes">
goes with audio06.mp3
</div>

## Local Search - Discrete Variables
\begin{minipage}{\textwidth}
\parbox{.6\textwidth}
{\includegraphics[width=.6\textwidth]{tsp1.pdf}}
\parbox{.4\textwidth}{
\begin{itemize}
\item Let $x_0 = $ 1-2-3-4-5-6-7-1, with cost 69
\item Swap cities 2 and 3 to get $x_1 = $ 1-3-2-4-5-6-7-1, with cost 68
\item Continue making swaps that improve the cost
\item Stop when no improving swaps are possible
\item No guarantee that result is best possible tour
\end{itemize}
}
\end{minipage}

<div class="notes">
goes with audio07.mp3
</div>

## Local or Global Minimum?

Minimize $f(x) = 0.5 x^5 - 6 x^4 + 24.5 x^3 - 39x^2 + 20 x$ subject to $0 \leq x \leq 5.$

\includegraphics[width=4in]{picture4.pdf}

<div class="notes">
goes with audio08.mp3
</div>

## Local vs Global again

\begin{minipage}{\textwidth}
\parbox{.5\textwidth}
{\includegraphics[width=.5\textwidth]{picture5.pdf}}
\parbox{.5\textwidth}{
{\includegraphics[width=.5\textwidth]{picture7.pdf}}
}
\end{minipage}

<div class="notes">
goes with audio09.mp3
</div>

## Local Search in R

For a smooth function (differentiable):

```{r eval=FALSE}
f <- function(x){
  0.5*x^5 - 6*x^4 + 24.5*x^3 - 39*x^2 + 20*x;
}
# find a local min starting from x = 2
x0 <- 2
optim(x0, f)
```

```{r echo=FALSE,warning=FALSE}
f <- function(x){
  0.5*x^5 - 6*x^4 + 24.5*x^3 - 39*x^2 + 20*x;
}
# find a local min starting from x = 2
x0 <- 2;
require(utils)
tmp <- noquote(
  capture.output(
    optim(x0, f) 
    )
  );
write.table(tmp[1:5],quote=F,row.names=F,col.names=F)
```

<div class="notes">
- goes with audio10.mp3
- For a one-dimensional function, that is, one with a single continuous input variable, using optim() in R
actually gives a warning to use optimize instead
</div>

## optimize() for 1D function in R

```{r}
f <- function(x){
  0.5*x^5 - 6*x^4 + 24.5*x^3 - 39*x^2 + 20*x;
}
optimize(f,c(0,5)) # specify the interval
```

optimize() is supposed to find *the* minimum of a continuous $f$ on $[a,b]$, but it's really just another local search.  It will find other minima if you choose a good interval.

<div class="notes">
goes with audio11.mp3
</div>

## Local search for higher dimensions

```{r eval=FALSE}
f3 <- function(x){ # 3 minima 2D function from earlier
  -.5*exp(-.7*( x[1]^2   +(x[2]-1)^2))
  -.7*exp(-.7*((x[1]-2)^2+(x[2]+1)^2))
  -   exp(-.7*((x[1]+2)^2+(x[2]+1)^2));
}
x0 <- c(1,1)
optim(x0,f3)
```

```{r echo=FALSE, warning=FALSE}
f3 <- function(x){ # 3 minima 2D function from earlier
  -.5*exp(-.7*( x[1]^2   +(x[2]-1)^2))
  -.7*exp(-.7*((x[1]-2)^2+(x[2]+1)^2))
  -   exp(-.7*((x[1]+2)^2+(x[2]+1)^2));
}
x0 <- c(1,1)
require(utils)
tmp <- noquote(
  capture.output(
    optim(x0, f3) 
    )
  );
write.table(tmp[1:5],quote=F,row.names=F,col.names=F)
```

<div class="notes">
goes with audio12.mp3
</div>

## optim() in package 'stats'

- optim() is really a gradient based search method
- if the gradient is not supplied, it is approximated numerically
- for a smooth function for which the gradient is not available, it is probably better to use a "derivative-free" method from package 'nloptr'
    - use newuoa() in 'nloptr' for unconstrained optimization
    - use bobyqa() in 'nloptr' for box constrained problems
    - use cobyla() for more complex constraints
- for large numbers of variables it really pays to get the gradient if possible, but the derivative-free algorithms can handle 100's of variables

<div class="notes">
goes with audio13.mp3
</div>

## nloptr package

- provides interface to open-source NLopt
- uniform syntax for many optimization algorithms
- can be a bit tricky to install (not required for homework)
    - some instructions in download packet (but Google is your friend)
    - for Windows get precompiled NLopt binaries and also Rtools
    - for Mac/Linux download NLopt source and compile
    - install.packages('nloptr') in R
    
<div class="notes">
goes with audio14.mp3
</div>

## newuoa() from nloptr

- local search, derivative-free method for unconstrained optimization
- many more options are possible, see documentation

```{r eval = FALSE}
library(nloptr); x0 <- c(1,1);
opts <- list("algorithm"="NLOPT_LN_NEWUOA", "xtol_rel"=1.0e-7);
result <- nloptr(x0=x0,eval_f=f3,opts=opts)
result
```

- output on next slide
- algorithm due to Michael Powell (in "Mathematical Programming", v. 100, p. 183-214, 2004)

<div class="notes">
put below slide:  there is no audio for this or the next three slides.  if you can install nloptr, then you may wish to try the code yourself.  The source code for these is embedded in the Week_09_Storybook.Rmd file in the download packet
</div>

## newuoa() output

- happens to find global optimum even though it doesn't start nearby

```{r echo = FALSE}
library(nloptr)
opts <- list("algorithm"="NLOPT_LN_NEWUOA", "xtol_rel"=1.0e-7);
result <- nloptr(x0=x0,eval_f=f3,opts=opts)
require(utils)
tmp <- noquote(
  capture.output(
    result
    )
  );
write.table(tmp[6:15],quote=F,row.names=F,col.names=F)
```

## bobyqa() from nloptr

- local search, derivative-free method for box constrained optimization
- many more options are possible, see documentation

```{r eval = FALSE}
library(nloptr)
opts <- list("algorithm"="NLOPT_LN_BOBYQA", "xtol_rel"=1.0e-7,
             lb=c(-2.5,-2.5),ub =c(2.5,2.5));
result <- nloptr(x0=x0,eval_f=f3,opts=opts)
result
```

- output on next slide
- algorithm due to Michael Powell (in "Mathematical Programming", v. 100, p. 183-214, 2004)

## bobyqa() output

- happens to find global optimum even though it doesn't start nearby

```{r echo = FALSE}
library(nloptr)
opts <- list("algorithm"="NLOPT_LN_BOBYQA", "xtol_rel"=1.0e-7);
result <- nloptr(x0=x0,eval_f=f3,opts=opts)
require(utils)
tmp <- noquote(
  capture.output(
    result
    )
  );
write.table(tmp[6:16],quote=F,row.names=F,col.names=F)
```


## Multistart Methods

<div class="notes">
- video01.mp4 in video folder
This slide is a placeholder for a movie that illustrates trying different starting points and exploring the resulting local minima, then accepting the smallest one as our optimal solution.  It may not be a global minimum.
</div>

## Naive Multistart

- Choose a random sample of starting points in the feasible domain.  
- Start a local search at each one.
- Return the coordinates and function value of the lowest minimum.

<div class="notes">
- explore in HW
</div>


## Naive Multistart in R

```{r}
bestmin <- 100000;
for (j in 1:10){
  x0 <- as.vector(runif(2,min=-2.5,max=2.5));
  result <- optim(x0,f3,method="L-BFGS-B",lower=c(-2.5,-2.5),upper=c(2.5,2.5))
  if (result$value<bestmin){ bestmin = result$value; bestx = result$par}
}
bestmin
bestx
```

<div class="notes">
goes with audio15.mp3
</div>

## Rastrigin Function

in 2D $f(x,y) = 20 + x^2 + y^2 - 10 \cos( 2 \pi x) - 10 \cos( 2 \pi y), -5.12 \leq x,y \leq 5.12$

\includegraphics[width=3in]{picture6.pdf}

<div class="notes">
goes with audio16.mp3
</div>


## Naive Multistart on Rastrigin


```{r eval = FALSE}
fr <- function(x){
  20 + x[1]^2 + x[2]^2 - 10*cos(2*pi*x[1]) - 10*cos(2*pi*x[2])
}
bestmin <- 100000; set.seed(126)
for (j in 1:50){
  x0 <- as.vector(runif(2,min=-5.12,max=5.12));
  result <- optim(x0,fr,method="L-BFGS-B",lower=c(-5.12,-5.12),upper=c(5.12,5.12))
  if (result$value<bestmin){ bestmin = result$value; bestx = result$par}
}
bestmin
bestx
```

- output on next slide

<div class="notes">
goes with audio17.mp3
</div>

## Naive Multistart on Rastrigin - output


```{r echo = FALSE}
fr <- function(x){
  20 + x[1]^2 + x[2]^2 - 10*cos(2*pi*x[1]) - 10*cos(2*pi*x[2])
}
bestmin <- 100000; set.seed(126)
for (j in 1:50){
  x0 <- as.vector(runif(2,min=-5.12,max=5.12));
  result <- optim(x0,fr,method="L-BFGS-B",lower=c(-5.12,-5.12),upper=c(5.12,5.12))
  if (result$value<bestmin){ bestmin = result$value; bestx = result$par}
}
```

```{r}
round(bestmin)
round(bestx,4)
```

<div class="notes">
goes with audio18.mp3
</div>


## MLSL

- MLSL = Multi-Level Single Linkage algorithm for global optimization (Kan and Timmer, "Stochastic global optimization methods", Mathematical Programming, vol.39, p. 27-78, 1987)
- "clustering" method to avoid repeated detection of the same minimum
    - randomly pick a small set of potential local search starting points
    - evaluate $f$ at these points
    - start a local search at a point if it is one of the lowest starting points and is not too close to a previous local search starting point
    - repeat three above steps until satisfied with result
- available in R package 'nloptr' (uses an approximate gradient local search)

<div class="notes">
goes with audio19.mp3
</div>

## MLSL applied to Rastigrin Function

```{r eval=FALSE}
require(nloptr); x0<-c(5,5);
result <- mlsl(x0,fr,lower=c(-5.12,-5.12),upper=c(5.12,5.12),nl.info=TRUE)
```

\vspace{-.25in}
```{r echo=FALSE}
require(nloptr); x0<-c(5,5);
require(utils)
tmp <- noquote(
  capture.output(
    result <- mlsl(x0,fr,lower=c(-5.12,-5.12),upper=c(5.12,5.12),nl.info=TRUE)
    )
  );
write.table(tmp[10:18],quote=F,row.names=F,col.names=F)
```

<div class="notes">
goes with audio20.mp3
</div>

## Genetic Algorithms

- objective function not smooth or very complicated
- finding "good enough" solutions in combinatorial optimization problems
- basic idea
    1. start with population of potential solutions (chromosomes)
    2. fittest individuals selected for breeding (selection)
    3. offspring formed by crossover and mutation
    4. offspring placed in new population and repeat
    5. if satisfied stop, else goto step 2.
    
<div class="notes">
- goes with audio21.mp3
PUT BELOW SLIDE: in addition to reading the textbook, there is a very good introduction at http://www.obitko.com/tutorials/genetic-algorithms/index.php
</div>

## Binary Encoding
  
- variables are represented as string of bits 
    - chromosome A: 100100101
    - chromosome B: 011010110
- Knapsack Problem
    - have a number of items of given value and size
    - knapsack with given capacity
    - select items to maximize value in knapsack 
    - encoding:  each bit represents whether an item is the knapsack or not

<div class="notes">
goes with audio22.mp3
</div>

## Permutation Encoding

- each chromosome is a permutation of the numbers $1, 2, ..., n$
- useful for ordering problems
- Traveling Salesman Problem
    - salesman visits all cities while minimizing cost (or distance)
    - chromosome represents order of visiting cities

<div class="notes">
goes with audio23.mp3
</div>

## Value Encoding

- each chromosome is string of some values such as reals, integers, characters
- real chromosomes can be used for complex optimization problems such as Rastrigin
- in the homework is an integer chromosome problem where integers represent assignment of cities to numbered congressional districts

<div class="notes">
goes with audio24.mp3
</div>

## Genetic Algorithms in R

- the 'GA' package is very good and can deal with binary, permutation, and real encoding
- the 'gramEvol' package has a useful genetic algorithm for integer encoded problems 
- both can be installed easily using `install.packages('name of package')`

<div class="notes">
there is no audio for this one
</div>

## GA for Knapsack Problem in R

```{r eval=FALSE}
value = c(4,6,5,3,6,3,7,8,9,10); 
size = c(3,2,5,3,2,4,3,2,4,2); capacity = 15;
totalValue = function(xb){
  sum(value*xb)*(sum(size*xb)<=capacity)
}
require(GA)
result = ga("binary",fitness=totalValue,nBits=10) # approx *max*
result@fitnessValue # total value of best knapsack
result@solution # binary vector showing which items to include
sum(size*result@solution) # capacity used
```

- output on next slide
- run this a few times, you won't always get the same solution!

<div class="notes">
goes with audio25.mp3
</div>

## Knapsack output

```{r echo=FALSE,warning=FALSE,message=FALSE}
value = c(4,6,5,3,6,3,7,8,9,10); 
size = c(3,2,5,3,2,4,3,2,4,2); capacity = 15;
totalValue = function(xb){
  sum(value*xb)*(sum(size*xb)<=capacity)
}
require(GA); require(utils);
result = ga("binary",fitness=totalValue,nBits=10);
    result@fitnessValue
    result@solution
    sum(size*result@solution)
```

<div class="notes">
no audio here
</div>

## GA for a TSP Problem in R

We'll use GA to approximate a solution to the sample TSP problem with 7 cities from an earlier slide.  Here is the cost matrix with a large cost for infeasible connections:

```{r eval=FALSE}
M = 100000;
costMatrix = as.matrix(rbind( 
                    c( 0,12,10, M, M, M,12),
                    c(12, 0, 8,12, M, M, M),
                    c(10, 8, 0,11, 3, M, 9),
                    c( M,12,11, 0,11,10, M),
                    c( M, M, 3,11, 0, 6, 7),
                    c( M, M, M,10, 6, 0, 9),
                    c(12, M, 9, M, 7, 9, 0)));
numcities = 7;
```

<div class="notes">
goes with audio26.mp3
</div>

## GA for TSP

Here is the fitness function:

```{r eval=FALSE}
# given a tour, calculate the total cost
tourCost <- function(tour, costMatrix) {
  tour <- c(tour, tour[1])
  route <- embed(tour, 2)[, 2:1]
  sum(costMatrix[route])
}
# inverse of the total distance is the fitness
tspFitness <- function(tour, ...) 1/tourCost(tour, ...)
```

<div class="notes">
goes with audio27.mp3
</div>

## GA for TSP

Run the GA:

```{r eval=FALSE}
result <- ga(type = "permutation", fitness = tspFitness, costMatrix=costMatrix, min = 1, 
             max = numcities, popSize = 10, maxiter = 500, run = 100, pmutation = 0.2            
             , monitor = NULL)
soln <- as.vector(result@solution[1,]) # use first soln
tourCost(soln,costMatrix)
tour <- c(soln,result@solution[1]);
tour # approx best tour
```

- for a large problem increase popSize and maxiter
- run=100 means the algorithm will terminate if there are 100 iterations without improvement
- pmutation = 0.2 is the probability of a mutation, default is 0.1
- output on next slide

<div class="notes">
goes with audio28.mp3
</div>

## GA for TSP

```{r echo=FALSE,warning=FALSE,message=FALSE}
M = 100000;
costMatrix = as.matrix(rbind( 
                    c( 0,12,10, M, M, M,12),
                    c(12, 0, 8,12, M, M, M),
                    c(10, 8, 0,11, 3, M, 9),
                    c( M,12,11, 0,11,10, M),
                    c( M, M, 3,11, 0, 6, 7),
                    c( M, M, M,10, 6, 0, 9),
                    c(12, M, 9, M, 7, 9, 0)));
numcities = 7;
# given a tour, calculate the total cost
tourCost <- function(tour, costMatrix) {
  tour <- c(tour, tour[1])
  route <- embed(tour, 2)[, 2:1]
  sum(costMatrix[route])
}
# inverse of thetotal distance is the fitness
tspFitness <- function(tour, ...) 1/tourCost(tour, ...);
set.seed(123) # make sure it works for demo, won't always find min
result <- ga(type = "permutation", fitness = tspFitness, costMatrix=costMatrix, min = 1, 
             max = numcities, popSize = 10, maxiter = 500, run = 100, pmutation = 0.2            
             , monitor = NULL);
soln <- as.vector(result@solution[1,]) # use first soln
"tour cost: "
tourCost(soln,costMatrix)
tour <- c(soln,result@solution[1]);
"approx best tour"
tour # approx best tour
```
- usually finds a tour equivalent to tour in textbook (p. 624)

<div class="notes">
goes with audio29.mp3
</div>

## GA for a larger TSP in R

\begin{minipage}{\textwidth}
\parbox{.5\textwidth}
{\includegraphics[width=.5\textwidth]{tour1.png}}
\parbox{.5\textwidth}{
{\includegraphics[width=.5\textwidth]{tour2.png}}
}
\end{minipage}

<div class="notes">
goes with audio30.mp3
</div>

## GA for the Rastrigin Problem

```{r eval=FALSE}        
Rastrigin <- function(x) {
            -(sum(x^2 - 10 * cos(2 * pi  * x)) + 10 * length(x))
        }
dimension = 10;
lower = rep(-5.12,dimension);
upper = rep(5.12,dimension);
result = ga(type="real-valued",fitness=Rastrigin,min=lower,max=upper,maxiter=500)
round(result@solution,5)
result@fitnessValue
```

<div class="notes">
goes with audio31.mp3
</div>

- results on next slide

## GA Rastrigin Results

- real-valued GA explores large solution space well, but converges slowly near optimum

```{r echo=FALSE}        
Rastrigin <- function(x) {
            -(sum(x^2 - 10 * cos(2 * pi  * x)) + 10 * length(x))
        }
dimension = 10;
lower = rep(-5.12,dimension);
upper = rep(5.12,dimension);
set.seed(123) # for reproducible results
result = ga(type="real-valued",fitness=Rastrigin,min=lower,max=upper,maxiter=500);
round(result@solution,5)
result@fitnessValue
```

<div class="notes">
goes with audio32.mp3
</div>

## GA plus local search

- add optim=TRUE to ga() call to use optim() to do local search using each chromosome as an initial value to determine fitness

```{r eval=FALSE}
result = ga(type="real-valued",fitness=Rastrigin,
            min=lower,max=upper,maxiter=1000,optim=TRUE)
```

```{r echo=FALSE}
set.seed(123)
result = ga(type="real-valued",fitness=Rastrigin,
            min=lower,max=upper,maxiter=1000,optim=TRUE);
round(result@solution,5)
result@fitnessValue
```

<div class="notes">
goes with audio33.mp3
</div>

## Simulated Annealing

- objective function not smooth or very complicated
- finding "good enough" solutions in combinatorial optimization problems
- basic idea for minimizing
    1. start with initial guess $x_o$ with fitness $f(x_n)$
    2. randomly choose a nearby point $x_n$ with fitness $f(x_n)$
    3. if $f(x_n) < f(x_o)$ then $x_o = x_n$ and goto step 2 if not converged
    4. if $f(x_n) \geq f(x_o)$ then $x_o = x_n$ with probability $P$ otherwise don't change $x_o$, goto step 2 if not converged
- probability $P$ of accepting worse solution decreases over time as simulation "cools"

<div class="notes">
goes with audio34.mp3

BELOW SLIDE: this video shows an animation of simulated annealing that I find really helpful in understanding how simulated annealing works.  Be sure to read the text below the video for an explanation.  (Include this link please)
https://youtu.be/iaq_Fpr4KZc
</div>

## Simulated Annealing in R

- simulated annealing is an option in optim() in package 'stats' 
- 'GenSA' package (check out the example of 30 dimensional Rastrigrin in documentation - hard problem!)

<div class="notes">
no audio here
</div>

## SA for TSP Problem in R

- see saTSP.R in download packet for the code 

\includegraphics[height=2.5in]{europe_tour.png}

<div class="notes">
goes with audio35.mp3
</div>

## SA for the Rastrigin Problem

```{r eval=FALSE}        
Rastrigin <- function(x) {
            (sum(x^2 - 10 * cos(2 * pi  * x)) + 10 * length(x))
        }
dimension = 10;
lower = rep(-5.12,dimension); upper = rep(5.12,dimension); 
x0 = runif(dimension,-5.12,5.12)
require(GenSA)
result = GenSA(x0,Rastrigin,lower=lower,upper=upper,
               control=list(max.call=5000))
result$value
result$par
```

- output on next slide

<div class="notes">
goes with audio36.mp3
</div>

## SA for Rastrigin output

```{r echo=FALSE}        
Rastrigin <- function(x) {
            (sum(x^2 - 10 * cos(2 * pi  * x)) + 10 * length(x))
        }
dimension = 10;
lower = rep(-5.12,dimension); upper = rep(5.12,dimension); 
x0 = runif(dimension,-5.12,5.12)
set.seed(123) # for reproducible results
require(GenSA)
result = GenSA(x0,Rastrigin,lower=lower,upper=upper,
               control=list(max.call=5000));
"approximate minimum value"
round(result$value,4)
"occurs at"
round(result$par,4)
```

<div class="notes">
no audio
</div>
